\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
Computer systems are used to store and protect sensitive data. To effectively
guard information, software running on these systems must conform to high
quality standards to establish trust in the correct functioning. The effort
invested by an attacker to obtain certain data directly depends on the
information value and the available resources. Therefore, systems handling
sensitive data must be developed with utter care and diligence.

Attaining high assurance for today's monolithic operating systems is
exceptionally hard since they tend to be large and include complex
functionality. This fact is highlighted by regular security updates provided by
operating system vendors.

In contrast, microkernels are well suited as a basis for systems with high
demands on robustness. They are by definition small, which is a precondition
for rigorous verification of correctness. Additionally, they lend themselves to
the construction of component-based systems where the incorrect behavior of one
partition does not impact the whole system.

A separation kernel (SK) is a specialized microkernel that provides an execution
environment for multiple components that can only communicate according to a
given policy and are otherwise isolated from each other. Hence, the isolation
also includes the limitation of potential side- and covert channels. SKs are
generally more static and smaller than microkernels, which minimizes the
possibility of kernel failure and should simplify the application of formal
verification techniques.

Recent addition of advanced hardware virtualization support for the Intel x86
architecture has the potential of greatly simplifying the implementation of a
separation kernel. This thesis presents a design of a separation kernel for the
Intel x86 architecture using the latest Intel hardware features. An open-source
prototype written in SPARK demonstrates the viability of the envisioned concept
and the application of SPARK's proof capabilities increases the assurance of the
correctness of the implementation.
