\chapter{Conclusion}
This chapter provides a summary of the contributions and an outlook on possible
future work.

\section{Contributions}
The main results of this work are the separation kernel design and prototype as
well as the accompanying proof artifacts. To our knowledge it is the first
publicly and freely available separation kernel for the Intel x86/IA-32e
architecture.

Use of the SPARK language and tools for the realization of the Muen kernel
provides the base for a high assurance implementation. Full absence of
runtime errors and some additional properties have been proven. Since the SPARK
tools are freely available, anyone can reproduce the proofs in their own
environment.

Focus on use of the latest Intel hardware virtualization features and emphasis
on a simple design have resulted in a small code size. The kernel is thus well
suited for review. It is a sound basis for further formal verification such as
the application of the theorem prover Isabelle/HOL via the HOL-SPARK
verification environment \cite{berghofer:OASIcs:2012:3587}.

An example system modeled after a realistic component-based use-case has been
implemented to demonstrate the viability of the design and the usability of the
kernel prototype. Multicore support makes better use of modern processors that
feature an increasing number of logical CPUs.

The complete source code, build environment, supporting tools and documentation
are provided, allowing independent review of the complete TCB.

\input{concl_future_work}
