\section{Policy}\label{sec:policy}
All aspects of a system using the Muen kernel must be specified in a policy XML
file. The policy is composed of the following main parts:

\begin{itemize}
	\item Hardware
	\item Kernel
	\item Binaries
	\item Subjects
	\item Scheduling
\end{itemize}

XML was chosen as a specification language since it is human-readable and can
be automatically verified against a schema. Furthermore, there is an existing
Ada library called XML/Ada TODO:Ref, which is opensource and freely available.

Since XML processing is not done by any trusted part of the system, the policy
contained in an XML file is transformed into SPARK source using the policy
compilation tool \texttt{skpolicy}. This process is described in detail in
section TODO:Ref.

Each of the main policy parts is presented in the following sections. Preceding
these descriptions is the specification of data types, which are the basis for
the subsequent definition of policy elements. The data types and elements map
directly to their corresponding XML schema definitions.

\subsection{Data types}
This section describes basic data types that are used in the specification of
the system policy. They are referenced in later chapters, illustrating different
parts of the policy.

\input{types.tex}

\subsection{Hardware}
\label{subsec:hardware}
\input{hardware.tex}

\subsection{Kernel}
\input{kernel.tex}

\subsection{Binaries}
\input{binary.tex}

\subsection{Subjects}\label{subsec:subjects}
\input{subject.tex}

\subsection{Scheduling}
\input{scheduling.tex}
